---
title: "Determination of Most Harmful Weather Events Across the US"
author: "Lee Mendoza"
date: "Saturday, April 11, 2015"
output: html_document
---

## Synopsis 
Harmful weather events are determined by analyzing National Weather Service data for weather events recorded from 1994 through 2011. Events most harmful with respect to population health (those resulting in deaths and injuries) are determined as are those having economic consequences (those resulting in property and crop damage). The top 10 most harmful events, across the US, are shown. The most harmful events for each state are  also determined and shown graphically. These results may be of use to government or municipal managers who might be responsible for planning and preparing for severe weather events.

## Data Processing  
Weather event data from the National Weather Service (NWS) from 1950 - 2011 were downloaded and prepared for analysis. The data were classified according to NWS weather event categories methods for downloading, 
```{r, echo=FALSE, message=FALSE}
## load the requred libraries
require(data.table)
require(dplyr)
require(stringr)
require(xlsx)
require(ggplot2)
require(maps)
require(grid)
require(gridExtra)
require(knitr)
```
```{r cache=TRUE}
################################ DATA PROCESSING ###############################
## the data have already been downloaded and should be in the current directory. If not,
## uncomment the following two lines and download the data:
##setInternet2(use = TRUE)
##download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "FStormData.bz2", mode = "wb")

## it's not necessary to decompress the file, R can read it directly. See this link...
## https://entropicevolution.wordpress.com/2014/01/24/learning-r-r-can-read-compressed-files/
dt =  read.table("FStormData.bz2", header = TRUE, fill = TRUE, stringsAsFactors = FALSE, sep = ",")
```
```{r cache=TRUE}
## convert dates to a useful format
dt$BGN_DATE = as.Date(dt$BGN_DATE, "%m/%d/%Y")
dt$END_DATE = as.Date(dt$END_DATE, "%m/%d/%Y")

## FILTERING ###############################
## it appears that event reporting increased substantially beginning in 1993.
## Since years before 1994 are inconsistent, events occurring before 1994 are ignored
dt = filter(dt, BGN_DATE >= "1994-01-01")

## let's also ignore events that caused no injuries, fatalities, property damage, and crop damage
harmful = filter(dt, INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)

## CLASSIFICATION ##########################
## there are 449 different levels of event type in the data, many of which do not match
## the event categories in the National Weather Service Instructions 10-1605
## <https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf>
## The reported events should be classified according to the National Weather Service (NWS) 
## categories listed in the documentation. We'll start by creating a NWS column 
## which indicates the new assigned NWS category
harmful$NWS = NA

## eventmap.xlsx contains a mapping of the reported events in the original data to the 
## event categories in the NWS Instructions 10-1605.
## The map was manually created by determining the closest match of the reported event in the data
## to the NWS category. In cases in which the event was listed as a combination (such as
## Hurricane/High Winds) the first reported event listed was used to classify the event, 
## on the assumption that the data collector listed the most significant event first.

## load the mapping file from the current folder
map = read.xlsx2("eventmap.xlsx", 1, stringsAsFactors = FALSE)

## loop through the mapping file, adding the NWS-mapped event type to the harmful data 
for (i in 1:nrow(map)){
    harmful$NWS[which(harmful$EVTYPE == map$EVTYPE[i])] = map$NWS[i]
}

## make the NWS category a factor
harmful$NWS = as.factor(harmful$NWS)
    
## prepare the data to calculate economic consequence
## the new PropMult column will be used as a numeric factor (based on the PROPDMGEXP column)
## with which to multiply the PROPDMG amount. The default value is 1
harmful$PropMult = 1
harmful$PropMult[which(harmful$PROPDMGEXP == 'b' | harmful$PROPDMGEXP == 'B')] = 1000000000
harmful$PropMult[which(harmful$PROPDMGEXP == 'm' | harmful$PROPDMGEXP == 'M')] = 1000000
harmful$PropMult[which(harmful$PROPDMGEXP == 'k' | harmful$PROPDMGEXP == 'K')] = 1000
harmful$PropMult[which(harmful$PROPDMGEXP == 'h' | harmful$PROPDMGEXP == 'H')] = 100
harmful$PropMult[which(harmful$PROPDMGEXP == '0') ] = 1
harmful$PropMult[which(harmful$PROPDMGEXP == '2') ] = 100
harmful$PropMult[which(harmful$PROPDMGEXP == '3') ] = 1000
harmful$PropMult[which(harmful$PROPDMGEXP == '4') ] = 10000
harmful$PropMult[which(harmful$PROPDMGEXP == '5') ] = 100000
harmful$PropMult[which(harmful$PROPDMGEXP == '6') ] = 1000000
harmful$PropMult[which(harmful$PROPDMGEXP == '7') ] = 10000000

## calculate the costs
harmful$PropCost = as.numeric(harmful$PROPDMG * harmful$PropMult)

## we need a similar transformation for the crop damage calculation. The CROPDMGEXP
## column has fewer unique values than the PROPDMGEXP column
harmful$CropMult = 1
harmful$CropMult[which(harmful$CROPDMGEXP == 'b' | harmful$CROPDMGEXP == 'B')] = 1000000000
harmful$CropMult[which(harmful$CROPDMGEXP == 'm' | harmful$CROPDMGEXP == 'M')] = 1000000
harmful$CropMult[which(harmful$CROPDMGEXP == 'k' | harmful$CROPDMGEXP == 'K')] = 1000
harmful$CropMult[which(harmful$CROPDMGEXP == 'h' | harmful$CROPDMGEXP == 'H')] = 100
harmful$CropMult[which(harmful$CROPDMGEXP == '0') ] = 1

## now calculate the costs
harmful$CropCost = as.numeric(harmful$CROPDMG * harmful$CropMult)

## for discussion purposes, summarize the overall damage by event type, over all areas
damage = harmful %>%
    group_by(NWS) %>%
    summarize(Fatalities = sum(FATALITIES), Injuries = sum(INJURIES), 
              Property = sum(PropCost)/1000000000, 
              Crops = sum(CropCost)/1000000000, 
              Count = n())

## determine the top 10 most harmful events in each category...
fatalities = damage %>% 
        arrange(desc(Fatalities)) %>% 
        select(NWS, Fatalities) %>% 
        top_n(10, Fatalities) %>% 
        rename("Event" = NWS)

injuries = damage %>% 
        arrange(desc(Injuries)) %>% 
        select(NWS, Injuries) %>% 
        top_n(10, Injuries) %>% 
        rename("Event" = NWS)

property = damage %>% 
        arrange(desc(Property)) %>% 
        top_n(10, Property) %>% 
        select(NWS, Property) %>% 
        rename("Event" = NWS, "Property Damage ($B)" = Property) 

crops = damage %>% 
        arrange(desc(Crops)) %>% 
        top_n(10, Crops) %>% 
        select(NWS, Crops) %>% 
        rename("Event" = NWS, "Crop Damage ($B)" = Crops) 

mostharmful = cbind(fatalities, injuries, property, crops)

## Now calculate the damage by event type for each state
damagebystate = harmful %>%
    group_by(STATE__, NWS) %>%
    summarize(Fatalities = sum(FATALITIES), 
              Injuries = sum(INJURIES), 
              Property = sum(PropCost), 
              Crops = sum(CropCost), 
              Count = n())

## pull in the pre-prepared Federal Information Processing Standard (FIPS) names here from
## https://www.census.gov/geo/reference/ansi_statetables.html. The codes from the states, 
## outlying areas, and individual minor outlying islands were combined to provide
## a usable list.  The data are read from the spreadsheet in the current folder
## and merge with the damagebystate data
fips = read.xlsx("statefips.xlsx", 1, stringsAsFactors = FALSE)

## merge the fips info with the damagebystate. This will give us standard state names to go
## with all the damage
damagebystate = merge(damagebystate, fips, by.x = 'STATE__', by.y = 'FIPS', all.x = TRUE)

## for each state, determine the event leading to the maximum number of deaths
maxdeaths = damagebystate %>%
            group_by(AreaName) %>%
            filter(Fatalities == max(Fatalities)) %>%
            arrange(AreaName) %>%
            select(STATE__, AreaName, NWS, Fatalities)
maxdeaths$AreaName = tolower(maxdeaths$AreaName)

## for each state, determine the event leading to the maximum number of injuries
maxinjuries = damagebystate %>%
            group_by(AreaName) %>%
            filter(Injuries == max(Injuries)) %>%
            arrange(AreaName) %>%
            select(STATE__, AreaName, NWS, Injuries)
maxinjuries$AreaName = tolower(maxinjuries$AreaName)

## for each state, determine the event leading to the maximum property damage
maxpropdamage = damagebystate %>%
    group_by(AreaName) %>%
    filter(Property == max(Property)) %>%
    arrange(AreaName) %>%
    select(STATE__, AreaName, NWS, Property)
maxpropdamage$AreaName = tolower(maxpropdamage$AreaName)

## for each state, determine the event leading to the maximum crop damage
maxcropdamage = damagebystate %>%
    group_by(AreaName) %>%
    filter(Crops == max(Crops)) %>%
    arrange(AreaName) %>%
    select(STATE__, AreaName, NWS, Crops)
maxcropdamage$AreaName = tolower(maxcropdamage$AreaName)

###################### end of data processing

```
## Results  
The rankings of the `r nrow(mostharmful)` most harmful weather events across the US are shown in Figure 1 below. As you can see, the event resulting in the most fatalities is `r mostharmful[1, 1]`, the most injuries is `r mostharmful[1, 3]`, the most property damage is `r mostharmful[1, 5]`, and the most crop damage is `r mostharmful[1, 7]`  

**Figure 1. The ten most harmful weather-related events by category, combined across states. Property and Crop Damage is shown in billions ($B) of dollars**
```{r kable, results='asis', echo=FALSE}
## alignment string
alignment = c('r', 'r', 'r', 'r', 'r', 'r', 'r', 'r')
kable(mostharmful, align = alignment, digits = 2)
```




```{r fig.width=12, fig.height=12, fig.align='center'}
################# Processing code to determine  
## load us map info
states_map = map_data("state")

##getPalette = colorRampPalette(brewer.pal(12, "Set3"))
##getPalette = colorRampPalette(c("red", "yellow", "white", "green", "blue"))
getPalette = colorRampPalette(c("#00007F", "blue", "#007FFF", 
                                "cyan", "green", "#7FFF7F", 
                                "yellow", "#FF7F00", "red", "#7F0000"))
palette = c("#4EC06C", "#F45AB2", "#A12102", "#1A303C",
            "#EAE7F5", "#E5AA7A", "#44246B", "#A09921",
            "#7CB3FA", "#E5F8A7", "#3FFDD0", "#60CB25",
            "#54182D", "#F63D70", "#F6A7F4", "#3C825A",
            "#AA6D1D", "#84FC87", "#093525", "#EBEBBB",
            "#A7151A")

## join the states mapping data with the death data
statesdeath = merge(states_map, maxdeaths, by.x = "region", by.y = "AreaName")
statesdeath$Event = as.factor(statesdeath$NWS)

p1 = ggplot()
p1 = p1 + theme(axis.ticks = element_blank(), 
                axis.text.x = element_blank(), axis.text.y = element_blank(), 
                axis.title.x = element_blank(), axis.title.y = element_blank(),
                panel.grid.minor=element_blank(), panel.grid.major=element_blank())
p1 = p1 + geom_polygon(data = statesdeath, aes(x=long, y=lat, group = group, fill = Event), color = "black") 
p1 = p1 + scale_fill_manual(values = getPalette(length(unique(statesdeath$Event))))
p1 = p1 + ggtitle("Event causing most fatalities per state")

## join the states data with the injury data
statesinjury = merge(states_map, maxinjuries, by.x = "region", by.y = "AreaName")
statesinjury$Event = as.factor(statesinjury$NWS)

p2 = ggplot()
p2 = p2 + theme(axis.ticks = element_blank(), 
                axis.text.x = element_blank(), axis.text.y = element_blank(), 
                axis.title.x = element_blank(), axis.title.y = element_blank(),
                panel.grid.minor=element_blank(), panel.grid.major=element_blank())
p2 = p2 + geom_polygon(data = statesinjury, aes(x=long, y=lat, group = group, fill = Event), color = "black") 
p2 = p2 + ggtitle("Event causing most injuries per state")

## plot the population health data on the same page
grid.arrange(p1, p2, ncol=1, nrow=2)
```
**Figure 2. Events most harmful to population health in each state. The top panel indicates the leading cause of weather-related deaths in each state, while the bottom panel indicates the leading cause of weather-related injuries.**  

```{r fig.width=12, fig.height=12, fig.align='center'}
## join the states data with the property data
statesproperty = merge(states_map, maxpropdamage, by.x = "region", by.y = "AreaName")
statesproperty$Event = as.factor(statesproperty$NWS)
p3 = ggplot()
p3 = p3 + theme(axis.ticks = element_blank(), 
                axis.text.x = element_blank(), axis.text.y = element_blank(), 
                axis.title.x = element_blank(), axis.title.y = element_blank(),
                panel.grid.minor=element_blank(), panel.grid.major=element_blank())
p3 = p3 + geom_polygon(data = statesproperty, aes(x=long, y=lat, group = group, fill = Event), color = "black") 
p3 = p3 + ggtitle("Event causing most property damage per state")

## join the states data with the injury data
statescrops = merge(states_map, maxcropdamage, by.x = "region", by.y = "AreaName")
statescrops$Event = as.factor(statescrops$NWS)

p4 = ggplot()
p4 = p4 + theme(axis.ticks = element_blank(), 
                axis.text.x = element_blank(), axis.text.y = element_blank(), 
                axis.title.x = element_blank(), axis.title.y = element_blank(),
                panel.grid.minor=element_blank(), panel.grid.major=element_blank())
p4 = p4 + geom_polygon(data = statescrops, aes(x=long, y=lat, group = group, fill = Event), color = "black") 
##p4 = p4 + scale_fill_manual(values = getPalette(length(unique(statescrops$Event))))
p4 = p4 + scale_fill_manual(values = palette)
p4 = p4 + ggtitle("Event causing most crop damage per state")
## plot the economic consequences data on the same page
grid.arrange(p3, p4, nrow=2, sub = textGrob("Events with greatest economic consequences"))
```
## Summary  

